---
title: "Workshop 8 TEMPLATE"
subtitle: "9 am"
author: "Stella Symonds"
date: "05/22/26"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here

drought_exp <- read_xlsx(
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
  sheet = "First Harvest")
```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

plant_colors <- c(
  "Lotus scoparius" = lotsco_col,
  "Eschscholzia californica" = esccal_col,
  "Penstemon centranthifolius" = pencen_col,
  "Salvia leucophylla" = salleu_col,
  "Stipa pulchra" = stipul_col,
  "Encelia californica" = enccal_col,
  "Grindelia camporum" = gricam_col
)

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# create new object from drought_exp
drought_exp_clean <- drought_exp |> 
  # making column names lowercase with underscores
  clean_names() |> 
  # renaming species to full names 
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # renaming water treatment 
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |> 
  # reordering species levels 
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  # reordering water treatment levels
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # filter for species, treatment, SLA, and weight columns 
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  

```{r}
#| label: checking-data

str(drought_exp_clean)
```


Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: ten-rows

slice_sample(drought_exp_clean, # data frame to display
             n = 10) # select ten rows 
```


# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 

```{r}
#| label: scatterplot

# base layer ggplot
ggplot(data = drought_exp_clean, # specify data frame
       mapping = aes(x = sla, # sla is x axis
                     y = total_g)) + # total weight is y axis
  # first layer: points
  geom_point() + 
  # rename axes with units 
  labs(x = "SLA (cm"^2~"/g)", # or (cm \U00B2 /g)
       y = "Total mass (g)",
       # add descriptive title 
       title = "as SLA increases, total mass increases")
```


## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.

```{r}
#| label: jitterplot-by-treatment

# base layer ggplot 
ggplot(data = drought_exp_clean,
       # assign axes
       mapping = aes(x = water_treatment,
                     y = total_g,
                     # color by treatment 
                     color = water_treatment)) + 
  # first layer: jitter
  geom_jitter(height = 0,
              width = 0.2) + 
  # custom colors
  scale_color_manual(values = c("Drought stressed" = ds_col,
                                "Well watered" = ww_col)) + 
  # rename axes and add title 
  labs(x = "Water treatment",
       y = "Total mass (g)",
       title = "Median well watered plant mass is\nhigher than drought stressed plant mass") + 
  # taking out legend 
  theme(legend.position = "none")
```


## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass.

```{r}
#| label: jitter-by-species

# base layer ggplot, assign axes
ggplot(data = drought_exp_clean,
       mapping = aes(x = species_name,
                     y = total_g,
                     # color by species 
                     color = species_name)) + 
  # first layer: jitter
  geom_jitter(height = 0,
              width = 0.2) + 
  # rename axes and add title 
  labs(x = "Species",
       y = "Total mass (g)",
       title = "Median plant mass between species") + 
  # custom colors 
  scale_color_manual(values = plant_colors) + 
  # take out legend 
  theme(legend.position = "none") + 
  # x axis labels don't overlap
  scale_x_discrete(labels = label_wrap(10))
```


# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model
model0 <- lm(
  total_g ~ 1, # formula: no predictors 
  data = drought_exp_clean # specify data frame
)

# model 1: all predictors
model1 <- lm(
  total_g ~ sla + water_treatment + species_name, # formula: all predictors
  data = drought_exp_clean # specify data frame
)

# model 2: SLA and water treatment
model2 <- lm(
  total_g ~ sla + water_treatment, # formula: SLA + water treatment
  data = drought_exp_clean # specify data frame
)

# model 3: SLA and species
model3 <- lm(
  total_g ~ sla + species_name, # formula: SLA + species
  data = drought_exp_clean # specify data frame
)

# model 4: water treatment and species
model4 <- lm(
  total_g ~ species_name + water_treatment, # formula: species + water treatment
  data = drought_exp_clean # specify data frame
)

# model 5: SLA
model5 <- lm(
  total_g ~ sla, # formula: only SLA
  data = drought_exp_clean # specify data frame
)

# model 6: water treatment
model6 <- lm(
  total_g ~ water_treatment, # formula: only water treatment
  data = drought_exp_clean # specify data frame
)

# model 7: species
model7 <- lm(
  total_g ~ species_name, # formula: only species
  data = drought_exp_clean # specify data frame
)
```

## Model diagnostics

```{r}
#| label: check-diagnostics

# ?
par(mfrow = c(2,2))
# generating diagnostic plots for best model (by AIC value)
plot(model4)
```

- residuals appear normally distributed and homoscedastic
- no clear outliers in data set influencing model coefficient estimates 


## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: calculating-aic

# calculating AIC
AICc(
  model0,
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7
) |> 
  # arranging output in order of AIC 
  arrange(AICc)
```

Insert code to display the summary for the best model. 

```{r}
#| label: model-summary

summary(model4)
```

- `(Intercept)`: model prediction for plant mass for Lotus Scoparius in the drought stressed treatment (doesn't show up in summary)
- `water_treatmentWell watered`: model prediction for difference in Lotus Scoparius mass between the well watered and drought stressed treatment 
- `species_namePenstemon centranthifolius`: model prediction for the difference between Penstemon mass in the well watered treatment and Lotus mass in well watered treatment 

## Stop and think: what does this model mean?  

**What is the best model?**

includes water treatment and species as predictors 

**How much variation in the response (total mass, in grams) does this model explain?**

72.6% of variation in plant mass is explained by a model with both of these predictors

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

- treatment influences plant mass
- species influences plant mass

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: model-predictions

preds <- ggpredict(
  model4, # specify model 
  terms = c("water_treatment", "species_name")) |> 
  # renaming columns 
  rename(water_treatment = x,
         species_name = group)
```


# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}
#| label: final-figure

ggplot(data = drought_exp_clean,
       mapping = aes(x = water_treatment,
                     y = total_g,
                     color = water_treatment)) + 
  # create separate panels for each species 
  facet_wrap(~ species_name) + 
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) + 
  geom_pointrange(data = preds,
                  mapping = aes(x = water_treatment,
                                y = predicted,
                                ymin = conf.low,
                                ymax = conf.high)) + 
  # take out legend 
  theme(legend.position = "none")
```


# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







